\section{Experiments}

[TODO: the final set of experiments is still being decided; the descriptions below outline candidate directions]

\subsection{Real World Tasks}

We plan to evaluate on the same real world manipulation tasks as SmolVLA, with SmolVLA, ACT, and a diffusion policy as baselines. We also want to test zero-shot generalization in the style of DreamZero on tasks not seen during fine tuning.

\subsection{LIBERO}

We plan to report results on LIBERO as a standardized benchmark, for more direct comparison against prior work.

\subsection{Ablations}

\subsubsection{World Modelling Objective}

Does the video latent prediction objective help, and when? We want to measure its effect in the single task and multi task regimes, with and without pretraining.

\subsubsection{JEPA vs Video VAE vs Pixel Space}

Does a V-JEPA2 backbone actually help, or would a similarly sized video VAE work just as well? Candidate drop-in replacements include the CogVideoX VAE, OmniTokenizer-VAE, and Cosmos-Tokenizer-CV. We also want to check whether pixel space diffusion is competitive at this scale.

\subsubsection{Pretraining}

Does pretraining on community data improve downstream task performance compared to training from scratch?

\subsubsection{Diverse Data}

Does fine tuning on diverse tasks lead to cross task generalization relative to single task fine tuning?

\subsection{Cross-Embodiment Transfer}

Can the model learn out of distribution tasks from video-only data on different embodiments, including egocentric human video?

\subsection{Classifier Free Guidance}

DreamZero uses CFG at inference. Does it help here, and does it change how the policy behaves?

\subsection{Continual Learning}

Can the video prediction component keep improving from the policy's own rollouts?
