\section{Method}

\subsection{Overview}

LeWAM jointly predicts future video latents and robot actions using a single conditional flow-matching Diffusion Transformer~\cite{DiT, FlowMatching, esser2024scaling}, following the joint video-action denoising approach of DreamZero~\cite{DreamZero}. Crutially, LeWAM learns to denoise video latents provided by a video encoder, not pixels~\cite{LeWorldModel}.

The model processes a sequence of three token groups: (1) \textbf{context video tokens} (clean), (2) \textbf{future video tokens} (noisy), and (3) \textbf{action tokens} (noisy), with a block-causal attention mask that enforces temporal ordering.
A frozen V-JEPA~2.1 encoder~\cite{vjepa2_1} provides the video representation used for both context and future ground truth. By predicting directly in latent space, the DiT backbone does not need to waste parameters encoding fine-grained, unpredicable, or irrelevant visual information. 

Additionally, a frozen SmolVLM2~\cite{SmolVLM2, SmolVLA} and an unfrozen state encoder provide visually-grounded language and state conditioning via cross-attention~\cite{GenTron}.

Figure~\ref{fig:arch} illustrates the full architecture.



%---------------------------------------

\subsection{Video Encoder}

We use V-JEPA~2.1~\citep{vjepa2_1} ViT-B as our visual encoder, initialized from pretrained weights and frozen during training. The ViT-B variant was distilled from the main ViT-G model, reducing it to only 80 million parameters. [TODO: can any evidence be foudn that it still works quite well despite such reduced size?]

All camera views are concatented horizontally, then divided into spatial patches of size $16 \times 16$ and grouped into tubelets of temporal size 2, halving the temporal token count.
We use a crop size of $224 \times 224$ per input image in practice, producing a $14 \times 14$ token grid per camera per frame, although the use of 3D-RoPE interpolation~\cite{ThreeDRoPE} allows for this value to be increased at run-time. [TODO: crop size should be increased with fine tuning, following the resolution scaling approach of Cosmos Predict~2.5~\cite{CosmoPredict} and V-JEPA~2~\cite{vjepa2_1}.]
Given $C$ context frames and horizon $H$, the encoder produces context latents $\mathbf{z}_{0:C} \in \mathbb{R}^{H_p W_p \times D_{\text{enc}}}$ and future targets $\mathbf{z}_{C:H} \in \mathbb{R}^{H_p W_p \times D_{\text{enc}}}$, where $D_{\text{enc}} = 768$.

%---------------------------------------

\subsection{Visually Grounded Language Conditioning}

We use SmolVLM2-256M~\cite{SmolVLM2, SmolVLA} as our language encoder for its small size and multi-image support. The VLM's vision encoder and connector are used to encode image context, which is prepended to the language token sequence before passing through the truncated language model layers. Inspired by ~\citet{CosmoPredict} and \citet{DecoderLLMTextToImage}, we extract hidden states from the first $N$ layers of the VLM's language model, concatenate them along the feature dimension, and project the result to the DiT's model dimension $D$ via a two-layer MLP. The entire VLM is frozen during training; only the projection MLP is learned.

This multi-layer feature extraction captures both low-level token semantics and higher-level compositional representations, providing richer conditioning than using the final layer alone~\cite{CosmoPredict}. Additionally, it allows us to reduce the depth of the VLM backbone, speeding up inference and training.

During training, language conditioning is dropped with probability $p_{\text{drop}} = 0.1$ to enable classifier-free guidance at inference time~\cite{DreamZero}.

%---------------------------------------

\subsection{Joint Flow-Matching DiT}

Unlike prior WAMs that use separate models for video prediction and action decoding~\cite{MimicVideo} [TODO: Cite more WAMs], we adopt the joint denoising formulation of ~\citet{DreamZero}, predicting both future video latents and actions jointly in a single DiT. The input sequence is constructed by concatenating the context, future, and action tokens.

All video tokens (context and future) are projected from $D_{\text{enc}} + 1$ to the model dimension $D$ via a shared linear layer, where the extra dimension is a binary flag specifying if a token is context ~\citep{CosmoPredict}.

\paragraph{DiT Block Structure.}
Each transformer layer consists of:
\begin{enumerate}
    \item \textbf{Block Causal Self-attention} over the full token sequence (see Appendix \ref{attn_structure} for masking details), with 3D-RoPE positional encoding~\cite{ThreeDRoPE, DreamZero, vjepa2_1}. Video tokens receive $(t, h, w)$ position IDs; action tokens receive $(t, 0, 0)$, sharing the temporal axis with their corresponding video blocks.
    \item \textbf{Cross-attention} to two external sources: (a) a proprioceptive state token, encoded from the robot's joint positions via a two-layer MLP, and (b) language tokens from SmolVLM2. The proprioceptive state provides the model with the robot's current configuration, grounding both video and action predictions in the physical state of the embodiment.
    \item \textbf{MLP} feedforward block.
\end{enumerate}
All three sub-layers use adaLN-Zero~\citep{DiT} conditioning on the flow-matching timestep embedding $\tau(t)$, allowing the network to modulate its behavior across the denoising trajectory. AdaLN weights are initialized to zero for stable early training.

\paragraph{Flow Matching Objective.}

Given a training sample with context latents $\mathbf{z}_{0:C}$, future targets $\mathbf{z}_{C:H}$, target actions $\mathbf{a}_{C:H}$, state $\mathbf{s}_{C-1}$, and language tokens $\mathbf{l}_{C-1}$, we sample noise $\boldsymbol{\epsilon}_z \sim \mathcal{N}(0, I)$, $\boldsymbol{\epsilon}_a \sim \mathcal{N}(0, I)$, and a shared timestep $t \sim \mathcal{U}[0, 1]$.
The noisy inputs are constructed via linear interpolation:
\begin{equation}
    \mathbf{z}_t = (1 - t)\,\boldsymbol{\epsilon}_z + t\,\mathbf{z}_{C:H}, \qquad \mathbf{a}_t = (1 - t)\,\boldsymbol{\epsilon}_a + t\,\mathbf{a}_{C:H}.
\end{equation}

The model predicts velocity fields $\hat{\mathbf{v}}_z, \hat{\mathbf{v}}_a = f_\theta(\mathbf{z}_t, \mathbf{a}_t, \mathbf{z}_{0:C}, t, \mathbf{s}_{C-1}, \mathbf{l}_{C-1})$, and the flow matching loss is:
\begin{equation}
    \mathcal{L}_{\text{flow}} = \left\| \hat{\mathbf{v}}_z - (\mathbf{z}_{C:H} - \boldsymbol{\epsilon}_z) \right\|^2 + \lambda_a \left\| \hat{\mathbf{v}}_a - (\mathbf{a}_{C:H} - \boldsymbol{\epsilon}_a) \right\|^2,
\end{equation}
where $\lambda_a$ balances the video and action losses.

\paragraph{Inference}

At inference time, we sample initial noise for both video latents and actions and integrate the learned velocity field from $t = 0$ to $t = 1$ using Euler integration:
\begin{equation}
    \hat{\mathbf{z}}_{t+\Delta t} = \hat{\mathbf{z}}_t + \hat{\mathbf{v}}_z \cdot \Delta t, \qquad \hat{\mathbf{a}}_{t+\Delta t} = \hat{\mathbf{a}}_t + \hat{\mathbf{v}}_a \cdot \Delta t.
\end{equation}

The predicted actions are optionally smoothed using a Savitzky-Golay filter to remove high frequency noise left over from the denoising process~\cite{DreamZero}.

\paragraph{Action normalization.}
Actions and proprioceptive state are normalized using precomputed dataset statistics. Values are clipped to the 2nd/98th percentiles and z-scored, preventing outliers from distorting the learning signal.

%---------------------------------------

\subsection{Training}

\paragraph{Robot training data.}
We train on the open-source HuggingFaceVLA community dataset utilized in SmolVLA~\cite{SmolVLA} (HuggingFace: \texttt{HuggingFaceVLA/community\_dataset\_v2}). The dataset is a collection of community-contributed robot manipulation demonstrations collected on the SO-101/100 low-cost arm, spanning 60 hours of in-the-wild data.

Most of the data preprocessing follows ~\citet{SmolVLA}, however new demonstrations were added in the second iteration of the dataset, prompting additional data preprocessing which was not done in the original SmolVLA project. Some added datasets did not match the SO101 emboidment, and a handful of demonstrations had undescriptive task labels; such episodes were either manually deleted or relabeled. 

Additionally, the original dataset was published using the LeRobotDataset v2.1 format, which is no longer supported. To better support modern tooling, we converted the entire to dataset to LeRobotDataset v3.0 and published it on HuggingFace (HuggingFace: \texttt{ehalicki/LeWAM\_community\_dataset})

\paragraph{Training schedule? Compute?}

%---------------------------------------

[TODO: add a python pseudocode, not the algorithmic package]
